A Plasma Torus Around a Young Low-Mass Star

### 2025-03-17 paper/ms.tex
update

\begin{abstract}
\normalfont
Approximately one percent of red dwarfs younger than 100 million years
show structured, periodic optical light curves suggestive of
transiting opaque material that corotates with the star
\cite{Rebull2016,Stauffer2017,Rebull2018,Bouma2024}.  The
composition, origin, and even the existence of this material are
uncertain. The main alternative hypothesis is that these complex
periodic variables (CPVs) are explained by complex distributions of
bright or dark regions on the stellar surfaces
\cite{Koen2021}.  Here, we present time-series spectroscopy and
photometry of a rapidly-rotating ($P$=3.9\,hr)
CPV, TIC~141146667. The spectra show coherent sinusoidal Balmer
emission at twice to four times the star's equatorial velocity,
demonstrating the presence of extended clumps of circumstellar
plasma --- a plasma torus.  Given that long-lived condensations of
cool ($10^4$ K) plasma can persist in the hot ($10^6$ K) coronae of
stars with a wide range of masses
\cite{CollierCameron1989,Townsend2005,Dunstone2006,Petit2013,Waugh2022,Daley-Yates2024},
these data support the idea that around the lowest-mass stars, such
condensations can be optically thick; the exact origin of the
material and its microphysical opacity remain unclear.
\end{abstract}

@@

{\it Radial Velocity}---We measured radial velocities of TIC~141146667
our HIRES spectra using a pipeline that we developed specifically for
rapidly rotating stars.  Our method is based on template-matching
against synthetic spectra produced by the PHOENIX stellar atmosphere
code \cite{Husser2013}.  We used the PHOENIX models with solar
metallicity and alpha element abundances, and calibrated our pipeline
using the standard stars described by \cite{Chubak2012}.  We used
velocity standards spanning spectral types from G2 to M4 (Barnard's
Star), irrespective of rotation rate.  We used \texttt{barycorrpy}
\cite{Kanodia2018} to calculate the velocity corrections due to Earth's
motion around the solar system barycenter and due to Earth's daily
rotation about its axis.  Our analysis code reproduces the systemic
velocities of known velocity standards \cite{Chubak2012} with an RMS of
0.66\,\kms.

For TIC~141146667, we measured the radial velocities using regions near
the K~I (7700\,\AA) resonance line and three TiO bandheads (5160\,\AA,
5450\,\AA, and 5600\,\AA).  We selected these regions because they
provided the best matches between the synthetic and observed spectra.
We then averaged the resulting redshift measurements over each order to
produce the final measurement.  We used the scatter of resulting
velocity measurements between orders to assign the RV uncertainty at
each epoch.  The uncertainty-weighted mean systemic velocity over all
epochs on 17~February~2024 (UT) was $\gamma$=0.6$\pm$1.5\,\kms.  The
relative radial velocities about this mean are given in
Table~\ref{tab:rv}.

@@

\begin{figure}[!t]
  \centering
  \includegraphics[width=0.7\textwidth]{figures/sf4.pdf}
  \caption{
    Spectral energy distribution of broadband photometric magnitudes
    plotted over the best-fit BT-Settl stellar atmosphere model
    \cite{Allard2012}.  This plot was made using \texttt{astroARIADNE}
    \cite{Vines2022}.  The photometry extends from the Gaia DR3 blue
    passband to WISE W3;  the redder (22 $\mu$m) W4 passband did not yield
    a confident detection.  This fit yields the star's temperature and
    size.  The lack of excess infrared flux relative to the photospheric
    model sets an upper limit on emission from any circumstellar dust.
    }
  \label{fig:sed}
\end{figure}

@@

{\it Upper and Low Bounds on Dust}---An upper limit on the amount of hot
dust follows from the lack of an infrared excess.  A lower limit follows
if one assumes that most of the broadband optical depth comes from
absorption and scattering by dust, rather than any radiative processes
associated with the plasma.

How much dust is allowed, given the lack of an infrared excess?
Figure~\ref{fig:sed} shows the SED.  While AllWISE \cite{Cutri2014}
yielded a confident W3 detection (9.8$\sigma$), the W4 extraction
yielded only a marginal (1.7$\sigma$) indication of detectable flux.
Similar to other CPVs \cite{Stauffer2017,Bouma2024}, the photometric
uncertainties from WISE W1 and W2 allow at most a $\lesssim$2\% excess
at 3-5\,$\mu$m relative to the stellar photosphere, and a $\lesssim$5\%
excess at 10\,$\mu$m (W3).  A simple approximation is to assume $T_{\rm
dust}$=1500\,K, typical for dust near the star (see \cite{Zhan2019} for
discussion regarding dust sublimation).  We then treat emission from the
dust and the star as Planck functions, requiring $F_{\lambda,\star} > f
\cdot F_{\lambda,\mathrm{dust}}$, for $f$ the scaling factor set by the
WISE photometric uncertainty.  For this condition to hold, we find that
the bolometric flux ratio $F_{\rm dust}/F_{\star}$$<$6$\cdot$10$^{-3}$.
Scaling the associated luminosities ($L = \sigma_{\rm SB} T_{\rm eff}^4
/ (4\pi R^2)$) implies an emitting area for the dust $<$10\% of the
star's surface area.  Converting this constraint to a dust mass requires
an assumption regarding the grain properties; taking a grain density
$\rho$=3\,g\,cm$^{-3}$ typical for silicate grains, and a fixed size
$a$=1\,$\mu$m \cite{Tanimoto2020,Gunther2022,Koen2023} yields an upper
limit of $M_{\rm dust}\lesssim$10$^{17}$\,g.

By way of comparison, the lower limit follows 
%FIXME TODO finish!!

The associated

which implies a 
Since the effective temperatures

luminosity

Under this approximation,

, reporting $W4$=8.4 with no uncertainty.  
The WISE images, available through the IRSA archive, concur.

%WISE_RSR_W3 9.8380 0.0450
%W4 = 8.392 +/- nan
%TODO TODO LEFT OFF HERE
...

@@

\begin{table}
  \centering
  \begin{tabular}{lcr}
  %\setlength{\tabcolsep}{2pt}
  \hline 
  \hline 
  Time [JD$_{\rm UTC}$] & RV (\kms) & $\sigma_{\rm RV}$ (\kms) \\
  % from tables/tab_TIC141146667_rel_rv.tex
  \hline 
  60357.450329 & 1.72 & 5.86 \\
  60357.461255 & -5.41 & 2.37 \\
  60357.472181 & -1.21 & 2.64 \\
  60357.483109 & 2.83 & 2.87 \\
  60357.494030 & 6.52 & 7.53 \\
  60357.504949 & -3.0 & 1.44 \\
  60357.515873 & 0.01 & 1.21 \\
  60357.526794 & -0.38 & 7.03 \\
  60357.537717 & -3.92 & 2.71 \\
  60357.548639 & 7.92 & 6.75 \\
  60357.559566 & 4.94 & 8.84 \\
  60357.570487 & -3.26 & 3.06 \\
  60357.581408 & 0.83 & 1.34 \\
  60357.592330 & 1.4 & 8.24 \\
  60357.603251 & -8.05 & 3.94 \\
  60357.614172 & -3.26 & 3.07 \\
  60357.625095 & -3.84 & 7.55 \\
  60357.636019 & -1.6 & 2.26 \\
  60357.646940 & 0.82 & 2.91 \\
  60357.657861 & 3.53 & 3.95 \\
  60357.668781 & 5.2 & 12.14 \\
  \hline
  \end{tabular}
  \caption{TIC~141146667 radial velocities.}
  \label{tab:rv}
\end{table}

>>>>

\begin{abstract}
\normalfont
Roughly one percent of red dwarfs younger than 100 million years
show structured, periodic optical light curves suggestive of
transiting opaque material that corotates with the star
\cite{Rebull2016,Stauffer2017,Rebull2018,Bouma2024}.  The
composition, origin, and even the existence of this material are
uncertain. The main alternative hypothesis is that these complex
periodic variables (CPVs) are explained by complex distributions of
bright or dark regions on the stellar surfaces
\cite{Koen2021}.  Here, we present time-series spectroscopy and
photometry of a rapidly-rotating ($P$=3.9\,hr)
CPV, TIC~141146667. The spectra show coherent sinusoidal Balmer
emission at twice to four times the star's equatorial velocity,
demonstrating the existence of extended clumps of circumstellar
plasma --- a plasma torus.  Given that long-lived condensations of
cool ($10^4$ K) plasma can persist in the hot ($10^6$ K) coronae of
stars with a wide range of masses
\cite{CollierCameron1989,Townsend2005,Dunstone2006,Petit2013,Waugh2022,Daley-Yates2024},
these data support the idea that around the lowest-mass stars, such
condensations can be optically thick; the exact origin of the
material and its microphysical opacity remain unclear.
\end{abstract}

@@

{\it Radial Velocity}---We measured radial velocities of TIC~141146667
from our HIRES spectra using a pipeline that we developed for
rapidly rotating stars.  Our method is based on template-matching
against synthetic spectra produced by the PHOENIX stellar atmosphere
code \cite{Husser2013}.  We used the PHOENIX models with solar
metallicity and alpha element abundances, and calibrated our pipeline
using the standard stars described by \cite{Chubak2012}.  We used
velocity standards spanning spectral types from G2 to M4 (Barnard's
Star), irrespective of rotation rate.  We used \texttt{barycorrpy}
\cite{Kanodia2018} to calculate the velocity corrections due to Earth's
motion around the solar system barycenter and due to Earth's daily
rotation about its axis.  Our analysis code reproduces the systemic
velocities of known velocity standards \cite{Chubak2012} with an RMS of
0.66\,\kms.

For TIC~141146667, we measured the radial velocities using regions near
the K~I (7700\,\AA) resonance line and three TiO bandheads (5160\,\AA,
5450\,\AA, and 5600\,\AA).  We selected these regions because they
provided the best matches between the synthetic and observed spectra.
We then averaged the resulting redshift measurements over each order to
produce the final measurement.  We used the scatter of resulting
velocity measurements between orders to assign the RV uncertainty at
each epoch.  The uncertainty-weighted mean systemic velocity over all
epochs on 17~February~2024 was $\gamma$=0.6$\pm$1.5\,\kms.  The
relative radial velocities about this mean are given in
Table~\ref{tab:rv}.

@@

\begin{figure}[!t]
  \centering
  \includegraphics[width=0.7\textwidth]{figures/sf4.pdf}
  \caption{
    Spectral energy distribution of broadband photometric magnitudes
    plotted over the best-fit BT-Settl stellar atmosphere model
    \cite{Allard2012}.  This plot was made from an adaptation of \texttt{astroARIADNE}
    \cite{Vines2022}.  The photometry extends from the Gaia DR3 blue
    passband to WISE W3;  the W4 passband (22 $\mu$m) did not yield
    a confident detection.  This fit yields the star's temperature and
    size.  The lack of excess infrared flux relative to the photospheric
    model sets an upper limit on emission from circumstellar dust.
    }
  \label{fig:sed}
\end{figure}

@@

{\it Upper and Low Bounds on Dust}---An upper limit on the amount of hot
dust follows from the lack of an infrared excess.  A lower limit follows
if one assumes that most of the broadband optical depth comes from
dust absorption and scattering, rather than any radiative processes
associated with the plasma.

How much dust is allowed, given the lack of an infrared excess?
Figure~\ref{fig:sed} shows the SED.  While AllWISE \cite{Cutri2014}
yielded a confident W3 detection (9.8$\sigma$), the W4 extraction
yielded only a marginal (1.7$\sigma$) indication of detectable flux.
Similar to other CPVs \cite{Stauffer2017,Bouma2024}, the photometric
uncertainties from WISE W1 and W2 allow at most a $\lesssim$2\% excess
at 3-5\,$\mu$m relative to the stellar photosphere, and a $\lesssim$5\%
excess at 10\,$\mu$m (W3).  To estimate the implied mass bound, we
assume $T_{\rm dust}$=1500\,K, typical for dust near the star (see
\cite{Zhan2019} for discussion regarding dust sublimation).  We then
treat emission from the dust and the star as Planck functions,
requiring $F_{\lambda,\star} > f \cdot F_{\lambda,\mathrm{dust}}$, for
$f$ the scaling factor set by the WISE photometric uncertainty.  For
this condition to hold, we find that the bolometric flux ratio $F_{\rm
dust}/F_{\star}$$<$6$\cdot$10$^{-3}$.  Scaling the associated
luminosities ($L = \sigma_{\rm SB} T_{\rm eff}^4 / (4\pi R^2)$)
implies an emitting area for the dust $<$10\% of the star's surface
area.  Converting this constraint to a dust mass requires an
assumption regarding the grain properties; assuming a grain density
$\rho_{\rm d}$=3\,g\,cm$^{-3}$ typical for silicate grains, and a fixed grain
size $a$=1\,$\mu$m \cite{Tanimoto2020,Gunther2022,Koen2023} yields an
upper limit of $M_{\rm dust}\lesssim$10$^{17}$\,g.

By way of comparison, the lower limit follows from requiring the
optical depth $\tau$ to be at least unity.
The optical depth can be written $\tau = n \sigma \ell = N \sigma$,
where $\sigma$ is the cross-section, $n$ is the number density, $\ell$
is the path length, and $N$ is the column density.
For spherical dust grains in the optical, $\sigma = Q_{\rm ext} \pi
a^2$, where $Q_{\rm ext}$ is the extinction efficiency parameter,
tabulated e.g.~by \cite{Croll2014} in their Figure 13.
\cite{Sanderson2023} calculated the relevant cloud mass scaling
assuming a spherical dust clump of size $r$,
and they found
\begin{equation}
  M_{\rm dust} \gtrsim 1.6 \cdot 10^{15}\, {\rm g}\ 
  \left( \frac{\tau}{1} \right)
  \left( \frac{Q_{\rm ext}}{3} \right)^{-1}
  \left( \frac{r}{0.1\,R_\star} \frac{R_\star}{0.34\,R_\odot} \right)^2
  \left( \frac{a}{1\,\mathrm{\mu m}} \right)
  \left( \frac{\rho_{\rm d}}{3\,\mathrm{g\,cm^{-3}}} \right).
\end{equation}
By way of comparison, prominences of the Sun typically have gas masses
of order $10^{14}\,g$-$10^{16}$\,g \cite{VialEngvold2015}, whereas a
planetesimal of this mass with a bulk density of 1\,g\,cm$^{-3}$ would
have a diameter of order one kilometer.  Halley's comet has a mass of
order $10^{17}$\,g \cite{Rickman1989}, of which of order one part in a
thousand is shed per orbit, most of which inspirals toward the Sun due
to Poynting-Robertson drag.

Broadly, these calculations suggest that if dust is responsible for
the broadband variability of CPVs, it would need to be concentrated in
clumps with masses in the range of $10^{15}$-$10^{17}$\,g.

@@

\begin{table}
  \centering
  \begin{tabular}{lcr}
  %\setlength{\tabcolsep}{2pt}
  \hline 
  \hline 
  Time [JD$_{\rm UTC}$] & RV (\kms) & $\sigma_{\rm RV}$ (\kms) \\
  % from tables/tab_TIC141146667_rel_rv.tex
  \hline 
  60357.450329 & 2.73 & 5.86 \\
  60357.461255 & -4.40 & 2.37 \\
  60357.472181 & -0.19 & 2.64 \\
  60357.483109 & 3.84 & 2.87 \\
  60357.494030 & 7.53 & 7.53 \\
  60357.504949 & -1.98 & 1.44 \\
  60357.515873 & 1.02 & 1.21 \\
  60357.526794 & 0.64 & 7.03 \\
  60357.537717 & -2.91 & 2.71 \\
  60357.548639 & 8.93 & 6.75 \\
  60357.559566 & 5.95 & 8.84 \\
  60357.570487 & -2.25 & 3.06 \\
  60357.581408 & 1.84 & 1.34 \\
  60357.592330 & 2.41 & 8.24 \\
  60357.603251 & -7.04 & 3.94 \\
  60357.614172 & -2.24 & 3.07 \\
  60357.625095 & -2.83 & 7.55 \\
  60357.636019 & -0.59 & 2.26 \\
  60357.646940 & 1.84 & 2.91 \\
  60357.657861 & 4.54 & 3.95 \\
  60357.668781 & 6.21 & 12.14 \\
  \hline
  \end{tabular}
  \caption{TIC~141146667 radial velocities relative to the systemic velocity.}
  \label{tab:rv}
\end{table}
